---
title: "The President's Covid-19 Network"
author: "Ansgar Wolsing"
date: "Oct 6th, 2020"
output: 
  html_document:
    bootstrap_version: 4+3
    theme: yeti
---

## Background

On Friday, Oct 2nd, ~~alleged tax evader~~ ~~TV celebrity~~ US President Donald Trump announced he got tested positive for Covid-19. In the days leading up to this announcement, Trump met lots of people ignoring any guidelines how to prevent spreading the virus. So let's revisit whom ~~the Superspreader in Chief~~ Trump encountered in the week before he announced he had contracted Covid-19.


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5.3,
  fig.process = "CairoPNG",
  dpi = 320,
  fig.retina = TRUE
)

```

```{r packages}

library(tidyverse)
library(lubridate)
library(ggtext)
library(ggraph)
library(tidygraph)

source("02_graphs_setup.R", verbose = FALSE)

```

## Scrape information

The New York Times has published an [article](https://www.nytimes.com/interactive/2020/10/02/us/politics/trump-contact-tracing-covid.html) with great visualization of the events Donald Trump participated before his positive Covid-19 test result was announced.

Let's scrape Trump's contacts at events from the article using `rvest` package and put everything into a tibble holding details about the meetings.


```{r scrape_data, cache=TRUE}
library(rvest)

# read the HTML content of the NYT article
url <- "https://www.nytimes.com/interactive/2020/10/02/us/politics/trump-contact-tracing-covid.html"
page <- read_html(url)

# extract days nodes
days <- html_nodes(page, css = ".g-day")

# extract the dates
day_names <- html_nodes(days, css = ".g-daySubHead") %>% 
  html_text()

# get the number of meetings for each day
meetings_each_day <- map(days, ~html_nodes(.x, css = ".g-meeting")) %>% 
  map_dbl(length)
names(meetings_each_day) <- day_names

# expand the day names vector by the number of meetings for each day, 
# so that 1 meeting is 1 row
day_names_expanded <- rep(day_names, meetings_each_day)

# extract meetings
meeting_nodes <- html_nodes(days, css = ".g-meeting")

# extract meeting names
meeting_names <- map_chr(meeting_nodes, ~html_nodes(.x, css = ".g-meetingSubhead") %>% 
                           html_text())

# tibble of meetings with meeting name and date to begin with
meetings <- tibble(meeting = meeting_names,
                   date_str = day_names_expanded)

meetings <- meetings %>% 
  mutate(wday = str_match(date_str, "(\\w+),")[, 2],
         date = str_match(date_str, ",\\s(.+)")[, 2])

# extract information for persons attending meetings
extract_contacts <- function(meeting) {
  person_nodes <- html_nodes(meeting, css = ".g-personStatus")
  person_names <- html_nodes(person_nodes, css = ".g-personName") %>% 
    html_text()
  person_covid_status <- html_nodes(person_nodes, css = ".g-personMug") %>% 
    html_attr("class") %>% 
    str_extract("test-(positive|negative)")
  
  tibble(person_name = person_names, person_covid_status)
}

# apply extraction to all person lists
meetings_persons <- map(meeting_nodes, extract_contacts)

# add persons to meetings tibble
meetings <- meetings %>% 
  bind_cols(tibble(persons = meetings_persons))

# clean up person data
meetings <- meetings %>%
  # unnest(persons) %>%
  mutate(persons = map(
    persons,
    ~ mutate(.x, 
      # cleanup person names by adding whitespaces where br tags were in the HTML
      person_name = str_replace_all(person_name, 
                                    "([a-z]|\\.)([A-Z])", 
                                    "\\1 \\2"),
      # cleanup covid status
      person_covid_status = case_when(
        person_covid_status == "test-positive" ~ "positive",
        person_covid_status == "test-negative" ~ "negative",
        is.na(person_covid_status) ~ "unknown"
      ) %>% as_factor()
    )
  ))

# store results with a timestamp in the filename
ts_str <- format(Sys.time(), "%Y%m%d-%H%M%S")
write_rds(meetings, 
          glue::glue("../output/trump_covid_tracing/trump_covid_tracing_meetings_{ts_str}"))

```



## Prepare the graph

Before building a graph of all contacts of Donald Trump prior to his Covid-19 announcement, we create all unique combinations of persons in the persons tibble for each meeting. The `combn` function from the `utils` package comes in handy. `combn` generates combinatons of the elements of a vector taken `m` elements at a time. The output is matrix (with `simplify = TRUE`), which has to be transposed with `t`. I wrap the call of `combn` in a mapper function which is than applied to each element of the persons lists.

```{r prep_graph}

# remove meetings with no persons listed
meetings <- meetings %>% 
  filter(map(persons, nrow) > 0)

# Build a tibble with all combinations of persons
build_connections <- function(x) {
  combn(x$person_name, 2) %>% 
    t() %>% 
    as_tibble() %>% 
    rename(person1 = V1, person2 = V2)
}

# map(meetings$persons, afunc)
meetings <- meetings %>% 
  mutate(person_connections = map(persons, build_connections))

# Covid status for all persons consolidated
person_covid_status <- meetings$persons %>% 
  bind_rows() %>% 
  distinct()

# Weighted connections
connections <- meetings$person_connections %>% 
  bind_rows() %>% 
  count(person1, person2, name = "weight", sort = TRUE)

# Attendents Rose Garden meeting
rose_garden_attendants <- meetings %>% 
  filter(meeting == "Rose Garden nomination ceremony for Amy Coney Barrett") %>% 
  unnest(persons) %>% 
  pull(person_name)

# create undirected graph from connections
graph <- as_tbl_graph(connections, directed = FALSE) %>% 
  activate(nodes) %>% 
  inner_join(person_covid_status, by = c("name" = "person_name")) %>% 
  mutate(rose_garden_attendant = (name %in% rose_garden_attendants)) %>% 
  mutate(centrality = centrality_degree())

```


## Meetings

Let's take a look how many people were tested positive on Covid-19 after having attended the meetings. Turns out many people attended the Rose Garden nomination ceremony for Amy Coney Barrett.

```{r meetings_plot, fig.height = 7.5}

# colors for Covid test results
colors <- RColorBrewer::brewer.pal(3, "Accent")
names(colors) <- c("negative", "positive", "unknown")

plot_subtitle <- sprintf("<b style='color:%s'>Positive</b>, <b style='color:%s'>negative</b>, or <b style='color:%s'>unknown</b> test results",
                          colors["positive"], colors["negative"], colors["unknown"])

meetings %>% 
  select(meeting, persons, date) %>% 
  unnest(persons) %>% 
  ggplot(aes(meeting)) +
  geom_bar(aes(fill = fct_rev(person_covid_status)), 
           width = 0.7,
           position = "stack",
           show.legend = FALSE) +
  scale_fill_manual(values = colors) +
  coord_flip() +
  facet_grid(rows = vars(date), scales = "free_y", space = "free") +
  labs(title = "ACB nomination a super-spreading event?",
       subtitle = plot_subtitle,
       x = NULL, y = "# participants")
  
```



## Network of contacts visualized

Let's plot the network of contacts and highlight all attendants of the Rose Garden nomination ceremony.

```{r network_vis, fig.width=8, fig.height=8}

plot_subtitle <- sprintf("<b style='color:%s'>Positive</b>, <b style='color:%s'>negative</b>, or <b style='color:%s'>unknown</b> test results<br>
                         The size of the edges indicates how many times people met when Trump was present.<br>
                         The bigger the nodes, the more contacts the person had with other persons.",
                          colors["positive"], colors["negative"], colors["unknown"])

set.seed(1234)
graph %>% 
  ggraph(layout = "stress") +
  geom_edge_arc(aes(width = weight, edge_alpha = weight), edge_color = "grey30",
                 show.legend = FALSE) +
  geom_node_point(aes(size = centrality, 
                      fill = person_covid_status, 
                      shape = rose_garden_attendant), 
                  alpha = 0.6,
                 show.legend = FALSE) +
  geom_node_text(aes(label = name), repel = TRUE, 
                 family = "Open Sans SemiBold",
                 size = 3,
                 show.legend = FALSE
                 ) +
  annotate("text", x  = -1.5, y = 1.5,
           hjust = 0, col = "grey25", family = "Inconsolata",
           label = "Squares indicate persons attended\nthe Rose Garden event",
           size = 3) +
  geom_curve(aes(x = -1.2, xend = -1, y = 1.35, yend = 1.15),
             curvature = 0.3, col = "grey50", size = 0.2,
             arrow = arrow(type = "closed", length = unit(1, "mm"))) +
  scale_edge_width(range = c(0.25, 1)) +
  scale_size(range = c(2, 12)) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = c("TRUE" = 22, "FALSE" = 21)) +
  labs(title = "Whom Trump met\nbefore his positive test on Covid-19 was announced",
       subtitle = plot_subtitle
       ) +
  # theme_bw() +
  theme(legend.position = "top",
        legend.justification = "left",
        plot.title = element_text(family = "Source Serif Pro SemiBold"))

```



***

Disclaimer: This exercise is completely unscientific. I'm  neither an epidemiologist nor virologist. Just some pet project scraping and data visualization. 

***



